A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay

### 2020-06-24 Paper/Science/main.tex
Submitted version

\documentclass[12pt,dvipsnames]{article}

%extra packages added during drafting, remove for submission
\usepackage{todonotes}
\usepackage{hyperref}
\usepackage{booktabs}

% Simple little macros to hide and highlight some text:
%\newcommand{\highlight}[1]{{\bf \textcolor{Maroon}{#1}}}
\newcommand{\highlight}[1]{{\bf {#1}}}

%\usepackage{ifthen}
\newif\ifshowoutline
%\showoutlinetrue
\showoutlinefalse
\ifshowoutline
    \newcommand{\outline}[1]{\textcolor{blue}{#1}}
\else
    \newcommand{\outline}[1]{}
\fi

\usepackage{amssymb}
\usepackage{xspace}

\gdef\arcsec{$^{\prime\prime}$}
%\def\SNABC{SN Requiem\xspace}

%\def\SNABC{AT 2016{\textcolor{red}{tbd}}\xspace}
%\def\reqlike{16{\textcolor{red}{tbd}}-like\xspace}

\def\SNABC{AT 2016{\textcolor{Gray}{tbd}}\xspace}
\def\reqlike{16{\textcolor{Gray}{tbd}}-like\xspace}
\def\lenstool{{\tt LENSTOOL}\xspace}

@@

%\title{Requiem for a 10 billion year old supernova } 
\title{A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay}
%\title{A Strongly Lensed Type Ia Supernova with a Decade-Long Time Delay}
%\title{SN Requiem: A Strongly Lensed Type Ia Supernova with a Decade-Long Time Delay}

@@

%%%%%%%fill in once decided
% \author
% {John Smith,$^{1\ast}$ Jane Doe,$^{1}$ Joe Scientist$^{2}$\\
% \\
% \normalsize{$^{1}$Department of Chemistry, University of Wherever,}\\
% \normalsize{An Unknown Address, Wherever, ST 00000, USA}\\
% \normalsize{$^{2}$Another Unknown Address, Palookaville, ST 99999, USA}\\
% \\
% \normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  jsmith@wherever.edu.}
% }

% Include the date command, but leave its argument blank.

\date{}

@@

@@

\begin{sciabstract}
%   The expansion history of the Universe is one of the most fundamental observations leading to the hot big bang model for the origin of the Universe\footnote{Penzias \& Wilson 1965, ApJL, 142, 419}.

  We report the discovery of a quadruply-lensed supernova (SN) explosion that will enable a time delay measurement with an uncertainty of $<1\%$. This unprecedented scenario will enable a rare and highly precise measurement of the Hubble constant from a single lensed SN.  The SN is in an evolved galaxy at $z=1.95$, gravitationally lensed by the massive foreground galaxy cluster MACSJ0138.  
 This is the most distant multiply-imaged supernova yet discovered, and we find it is classified as a Type Ia with {$>90\%$} probability.
  Three lensed images of the SN are detected in existing observations from the \textit{Hubble Space Telescope} with relative time delays of $<$200 days.  A fourth image in the component closest to the cluster core is predicted to appear in the year 2037$\pm$2.  Observing the light curve of that future image could provide a time delay precision of $\approx 7$ days over an extraordinary baseline of nearly 20 years.

@@

%\section*{Introduction}

%\todo[inline]{JP: Can edit intro} SN1a play a special role in cosmology, they are standard candles -> led to discovery of DE. High redshift SN-1a provide particularly strong constraints.
%\newline

@@

One promising method uses %time delays of 
gravitational lens systems in which a background
source appears as multiple images that arrive to the observer with relative delays \cite{einstein_lens_1936}.
%, which %for a given lens potential 
When such a strongly-lensed source is variable, one can measure the time delay between any pair of lensed images and derive a ratio of angular diameter distances to the foreground lens and background source. 
This distance ratio is sensitive to cosmological parameters, such as 
%the expansion rate of the universe, 
$H_0$  and the dark energy equation of state, $w$ \cite{refsdal_possibility_1964,coe_cosmological_2009,linder_lensing_2011}.
In recent years this method has been very successfully applied to lensed
quasars \cite{suyu_cosmology_2014,bonvin_cosmograil_2019}, with six high-precision measurements to date \cite{wong_h0licow_2019}. As this sample of time delay lenses grows to several dozen, it is expected to deliver a measurement of $H_0$ with 1\% precision \cite{suyu_cosmological_2018}.

Gravitationally lensed SN with multiple images present an attractive new addition to this field. Most notably, SN exhibit relatively simple photometric behavior, with well-understood light curve shapes and colors---a contrast to the stochastic variation of quasars.  However, the exploration of lensed SN time delay cosmography has been hindered by their rarity.  To date, there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_sn_2016}. The second, {\it SN 2016geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2019}. Here we present the discovery 
of a third lensed SN resolved into multiple images: \SNABC.

We discovered \SNABC using data from the Hubble Space Telescope (HST) program {\it REsolved QUIEscent Magnified Galaxies} (REQUIEM, HST-GO-15663, PI:Akhshik). This project targets rare examples of quiescent galaxies that have been magnified by strong gravitational lensing. %These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular galaxy targeted by REQUIEM
is MRG-M0138, first discovered as a massive red galaxy (MRG) at $z=1.95$ \cite{newman_resolving_2018} behind the massive galaxy cluster MACS J0138.0-2155 \cite{ebeling_macs_2001}.
MRG-M0138 is quadruply lensed by the foreground cluster, which is at $z=0.338$.  
During analysis of REQUIEM observations obtained 13--14 July 2019 \cite{materials_methods_2020} we discovered three point 
sources that were present in archival HST images from 18--19 July 2016, part of the program 
that first confirmed the MRG-M0138 galaxy as a strongly-lensed object (HST-GO-14496; PI:Newman). 
Each point source is within 5 arcseconds of one of the four MRG-M0138 images.  None of the
three point sources are present in the REQUIEM HST data in 2019 (Fig \ref{fig:layout}). We infer that 
the point sources are multiple images of a single astrophysical 
transient in MRG-M0138, most likely a SN \cite{materials_methods_2020}.  
%Hereafter we call this multiply-imaged transient \SNABC.
The official designation for this multiply-imaged astrophysical transient is \SNABC.

To construct a lens model for the MACS J0138.0-2155 cluster we used the publicly available \lenstool software \cite{jullo_bayesian_2007, kneib_lenstool_2011} to model the mass distribution in the cluster core as the combination of a cluster scale and multiple galaxy scale potentials \cite{materials_methods_2020}.  To avoid unintended bias, we kept the lens model development completely separate from the analysis of the SN.  Only upon independent completion of both were the results combined for the analysis described here.  The input model constraints are the positions and redshifts of the MRG-M0138 galaxy at $z=1.95$ (both the central light peak and the SN location in each image) as well as a multiply-imaged background 
galaxy at $z=0.766$, both having secure spectroscopic redshifts \cite{materials_methods_2020}.  
From this model we derive estimates for the lensing magnification and time delay of each of the SN images, including the predicted fourth image (Table \ref{tab:time_delays}).
The lensing model predicts that the SN should appear in the fourth MRG-M0138 image in the year 2037$\pm$1.5, demagnified with $\mu=0.4\pm0.2$. The predicted location for this future image is $\alpha=01:38:04.15,\delta=-21:55:24.659$.
A fifth image will also appear at a later date, located very near the center of the cluster and much more significantly demagnified, so therefore likely not observable.

%The range of possible transients for Requiem is broad, encompassing normal SNe, superluminous SNe (galyam luminous 2019), fast transients (berger fast 2013, drout rapidly 2014), and even gravitational microlensing events (rodney two 2018, kelly extreme 2018). 
To realize the cosmological promise of \SNABC 
we need to estimate the age of each SN image, which in turn constrains the lensing time delays. For this
goal, it is valuable to have a firm determination of the transient’s class. Expected time delays and magnifications from the lens model exclude any of the various rapidly evolving and low-luminosity stellar transient classes, strongly suggesting that it is a SN. The first-order SN distinction remaining is between a Type Ia SN---the explosion of a white dwarf star in a binary system---and a core collapse SN (CCSN)---the end-point of a star with mass $\gtrsim 10 M_{\odot}$. 
The properties of the host galaxy can inform this classification because CCSN are limited to galaxies with young stellar populations. Limits on the specific star formation rate and age for this host, MRG-M0138, show it to be a massive but very quiescent and evolved galaxy, %($\log(M_*/M_\odot)=11.7\pm0.2; )$ , 
unlikely to retain any significant population of high-mass stars \cite{newman_resolving_2018}. Based on observed properties of the host galaxy alone, we find a 62-75\% probability that \SNABC is a Type Ia SN \cite{materials_methods_2020}. Adopting the lens model magnifications for the three observed SN images (Table \ref{tab:time_delays}) the position of each SN image in color-magnitude space (Figure \ref{fig:class}) provides more classification leverage, yielding p(Ia)=95\% \cite{materials_methods_2020}. By also including the model-predicted time delays, we can treat the three SN images effectively as three points on a common SN light curve, and we find p(Ia)=94\% \cite{materials_methods_2020}.
%This is corroborated by the limited data available for the transient. In color -magnitude space it displays best agreement with the Type Ia population (Fig 1). 
An improved classification can be achieved with spectroscopy and multi-band photometry upon arrival of the fourth image.

%The galaxy has very low SFR=XX and based on stellar ages stopped forming stars ~X.X Gyr earlier. This is a proto typical early type galaxy progenitor.

%%%%%% ORIGINAL Classification text
% To evaluate the age of each transient image (and therefore constrain the lensing time delay between images), it is valuable to have a firm determination of the transient's class.  The range of possible transients is broad, encompassing normal SNe, superluminous SNe \citep{galyam_luminous_2019}, fast transients \citep{berger_fast_2013, drout_rapidly_2014}, and even gravitational microlensing events \citep{rodney_two_2018,kelly_extreme_2018}.  Identifying the transient type makes it possible to use the well-developed library of spectrophotometric SN models \citep[e.g.][]{kessler_models_2019} for the time delay measurement. 
%  The first-order distinction that is most important is between a Type Ia SN--understood to be the explosion of a white dwarf star in a binary system--and core collapse SN (CCSN)--the end-point of a star with mass $\gtrsim 10$M$_{\odot}$.  Given the limited data available for this transient, we will not attempt a more fine-grained classification, though in principle that could be achieved with spectroscopy and multi-band photometry upon arrival of the final image.

% As a first step, we can use inferences from the lens model to establish a strong prior against any of the various rapidly evolving and low-luminosity stellar transient classes.  
% The expected time delays between the images are $\sim100$ observer-frame days, 
% but we see that three images of the transient are visible simultaneously.
% From this we can infer that the visibility time of the transient in the $z=1.95$ rest-frame must be at least $\sim$30 days. 
% Similarly, with expected magnifications in the vicinity of $\mu\sim10$, the measured apparent magnitudes near 23 AB mag translate to a rest-frame absolute magnitude near $M_B \sim-19.5$ mag (too bright to be a nova, luminous blue variable, or other low-luminosity stellar transient).
% Taken together, these indicators strongly suggest that the transient is a supernova (SN).

% Although we have invoked the lens model in this analysis, note that the inferences are not strongly dependent on the specific lens model predictions.  To make the observed transient images consistent with a fast or low-luminosity transient, the time delays and/or magnifications would have to be changed by more than a factor of 2.  In the analysis to follow, we will work under the assumption that the transient is a SN.

% \subsection{Host galaxy properties}
% \label{ss:host}

% We first seek to classify this transient {\it circumstantially}, using measured properties of the host galaxy stellar population to infer the type. Although Type Ia SNe are found in all types of galaxies, CCSNe are limited to galaxies with relatively young stellar populations.  We can therefore infer some information about the SN type using the observed host properties combined with knowledge of the relative rates of Type Ia and CCSNe in different stellar populations \citep{mannucci_rates_2005,foley_classifying_2013}.  In the case of the host galaxy MRG0138, we have very stringent limits on the star formation rate. This galaxy is a high-mass but very quiescent galaxy, with a specific star formation rate of $\sim10^{-11.3}$ yr$^{-1}$  and a stellar population that is well-matched by an exponential star formation history with an age of $1.4$ Gyr \cite{newman_resolving_2018}.  The massive stars that end as CCSN explosions have main-sequence lifetimes of $\lesssim 40$ Myr,  making it highly unlikely that MRG0138 retains any significant population of massive stars \SR{this should be revised, b/c of total mass.  Focus instead on the relative probability. as discussed on telecon}

% Discussion from Newman et al. \citep{newman_resolving_2018-1}.

% Conclusion: the host is a red-dead elliptical. It is most likely a Type Ia SN.

% \subsection{Classification from Color and Brightness}
% \label{ss:lightcurve}

% Figure~\ref{fig:colormag} shows the location of \SNABC in color-magnitude space, comparing the de-magnified apparent F160W magnitude against the observed F105W-F160W color.  
% For a source at $z=1.95$ the F160W band is centered near 5400 \AA, approximately corresponding to the Bessell V bandpass, while the F105W-F160W color approximates a U-V color. 
% As shown in Figure~\ref{fig:colormag}, the observed photometry for \SNABC is fully consistent with the expected color and magnitude of a Type Ia SN at $z=1.95$, after correcting for lensing magnification using the $\mu$ values from Table~\ref{tab:model_evidence}.  The magnification-corrected photometry is inconsistent with Type II SNe, which are fainter and more blue. \SNABC is marginally consistent with the CCSN Type Ib/Ic sub-class, although the image 2 point falls outside of the 95\% confidence region for that population.

% The right panel in Figure~\ref{fig:colormag} shows that the three data points are also consistent with the expected time-dependence  of a Type Ia SN in this color-magnitude space. Visualizing the changing color and brightness of a SNIa at $z=1.95$ in the F105W and F160W bands, we see that it  would intersect image 2 near peak brightness and approach the image 1 and 3 photometry at roughly 80 observer-frame days after peak.

% Taken together, this evidence reinforces the conclusion that \SNABC is most likely a Type Ia SN.  Although we have to adopt a magnification correction 
% from a lens model, this conclusion is not sensitive to the choice of lens model.  Every lens model we have evaluated locates \SNABC's un-magnified position squarely within the Type Ia region of color-magnitude space.  
% Dust is also not a confounding factor here.  The simulations used to generate the contours in Figure~\ref{fig:colormag} include dust extinction at the source plane.  If a significant screen of dust exists in the lens plane, this would have the effect of making the SN appear dimmer and more red, so correcting for that extinction would move the \SNABC requiem points upward and leftward on this plot, causing it to be even more strongly associated with the Type Ia region.

% \SR{TODO: add an appendix explaining the sims}

The color of a SNIa evolves substantially over its lifetime as the photosphere expands and cools, revealing different layers of the expanding shell and driving episodes of recombination \cite{woosley_type_2007,kasen_type_2009}. Since the phenomenon of gravitational lensing in general is achromatic, this color evolution makes it possible to derive an age constraint that is largely independent of the lens model \cite{materials_methods_2020}.  Combining this information with magnification constraints from the lens modeling helps break parameter degeneracies, yielding the measured delays in Table \ref{tab:time_delays} \cite{materials_methods_2020}. 
Remarkably, the ages of image 1 and 3 are constrained to better than $\pm$20 days, despite having only a single epoch of photometric data.    These uncertainties may be 
further reduced when the future fourth image is observed with high-precision, multi-epoch photometry.  Such a light curve will pin down the intrinsic SN light curve parameters that are shared by all images, and break remaining parameter degeneracies (see supplemental materials).
%These constraints represent an upper limit on the age (and therefore time delay) uncertainties, but they would still deliver an uncertainty of $\sim0.3\%$ on the predicted $\sim6000$ day delay. While already dwarfed by the expected lens model uncertainty, constraints on the intrinsic SN light curve parameters gained from exquisite photometry of the future fourth image of Requiem could improve this even further.

Future large-scale surveys such as the Vera C. Rubin Observatory 
%Legacy Survey of Space and Time (LSST) 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their duration \cite{oguri_gravitationally_2010,goldstein_rates_2019,wojtak_magnified_2019}. However, many of these lensed SN will require substantial follow-up resources to be cosmologically useful. Furthermore, almost all will have significantly shorter 
delays (of order 10-100 days) and therefore higher relative uncertainties than \SNABC.  Since it is the {\it fractional} time delay uncertainty that 
propagates through to any time delay distance measurement, the extraordinarily long time delays of cluster-lensed SNe like SN Requiem can deliver significantly better time delay precision, with comparable observational cost.  In fact, the long time baseline of lensed SN like \SNABC effectively insures that their cosmological precision is not limited by the time delay measurement errors.

In Figure \ref{fig:cosmo} we illustrate the cosmological value of time delay cosmography with objects like \SNABC  by considering a small sample of 5 ``\reqlike'' SN.  For a comparison on equal footing, we evaluate equal-sized lensed SN samples 
that mimic those expected from the Rubin Observatory and  the Roman Space Telescope.  Differences in the 
ratio of source to lens redshift account for the different cosmological parameter covariances, reflected in the orientations of the contours in these plots. Because of the unusually large redshift ratio of \SNABC, the constraints 
%on dark energy equation of state parameters ($w_0$, $w_a$) 
from the \reqlike sample are nicely complementary to the low-$z$ Rubin Observatory and high-$z$ Roman Space Telescope samples. 
Furthermore, when we assume that the lens modeling uncertainty is equal for each sample (amounting to a 5\% uncertainty in each time delay distance), the \reqlike sample has leverage that is comparable to the same number of lensed SN from the wide-field surveys. However, if we improve  the lensing constraints to 2\% precision, the \reqlike contours tighten dramatically, reflecting the fact that these long-time-baseline events are limited only by the lens modeling uncertainty.

To achieve these cosmological constraints in practice, such a sample could be developed with regular monitoring of cluster-scale lenses, partnered with modest follow-up to characterize any lensed SN discovered.  As \SNABC shows, even a single epoch of photometric data at the time of detection may be sufficient to provide the necessary anchor point for this promising cosmological tool.  After that, all we need is patience.

@@

%BibTeX users: After compilation, comment out the following two lines and paste in
% the generated .bbl file. 
\clearpage
\begin{figure*}
    \centering
    \includegraphics[draft=False,width=0.9\textwidth]{Paper/Figures/fig1_layout_sn4.pdf}
    \caption{Overview of the MACSJ0138 cluster field and the locations of the \SNABC images  (SN1--3) and its multiply-imaged host galaxy (H1--4). The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  
    %Labels marked 3.1--3.4 indicate the locations of a separate  multiply-imaged [O\textsc{ii}]-emitter at $z=0.77$ used to help constrain the lensing potential. 
    Panels \emph{b}--\emph{i} show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels \emph{bcde} show the imaging from 2016 July where the SN was visible and \emph{fghi} show the later imaging from 2019 July where the the SN has faded away.  Panels \emph{ei} indicate the location of the fourth image predicted to appear in $\approx$2037. All panels use the late-epoch F125W imaging for the green channel; nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, helps to constrain the relative age of each SN image and the transient classification as a likely Type Ia supernova explosion.}

>>>>

\documentclass[12pt]{article}

%extra packages added during drafting, remove for submission
\usepackage{hyperref}
\usepackage{booktabs}

%\usepackage{amssymb}
\usepackage{xspace}
\usepackage{graphicx}

\gdef\arcsec{$^{\prime\prime}$}
\def\TNSname{AT 2016tbd\xspace}
\def\SNABC{AT 2016tbd\xspace}
%\def\SNABC{SN Requiem\xspace}
%\def\reqlike{Requiem-like\xspace}
\def\reqlike{16tbd-like\xspace}

\def\lenstool{{\tt LENSTOOL}\xspace}

@@

\title{A Gravitationally Lensed Supernova with an Observable Two-Decade Time Delay}

@@

\author{
 Steven A. Rodney$^{1\ast}$, 
 Gabriel Brammer$^{2\ast}$,  
 Justin D.\,R.~Pierel$^{1}$,\\ 
 Johan Richard$^{3}$, 
 Sune Toft$^{2}$,\\  
 Kyle F. O'Connor$^{1}$, 
 Mohammad Akhshik$^{4}$, 
 Katherine E. Whitaker$^{2,4,5}$  
\\
\normalsize{$^{1}$Department of Physics \& Astronomy, University of South Carolina,}\\ \normalsize{ Columbia, SC 29208, USA}\\
\normalsize{$^{2}$Cosmic Dawn Center, Niels Bohr Institute, University of Copenhagen,}\\ \normalsize{Lyngbyvej 2, Copenhagen \O\ 2100 Denmark}\\
\normalsize{$^{3}$Univ Lyon, Univ Lyon1, Ens de Lyon, CNRS, Centre de Recherche Astrophysique de Lyon}\\ \normalsize{UMR5574, F-69230, Saint-Genis-Laval, France}\\
\normalsize{$^{4}$Department of Physics, University of Connecticut,}\\ \normalsize{Storrs, CT 06269, USA}\\
\normalsize{$^{5}$Department of Astronomy, University of Massachusetts,}\\ \normalsize{Amherst, MA 01003, USA}\\
\normalsize{$^\ast$To whom correspondence should be addressed;}\\ \normalsize{E-mail:  srodney@sc.edu, gabriel.brammer@nbi.ku.dk}
}

% Include the date command, but leave its argument blank.
\date{}

@@

\paragraph*{One Sentence Summary:} A distant supernova has appeared as three images magnified by the gravity of a galaxy cluster, and will reappear in 20 years.

@@

\begin{sciabstract}

  We report the discovery of a strongly-lensed supernova (SN) explosion that will enable a time delay measurement with an uncertainty of $<1\%$. This unprecedented scenario will enable a rare and highly precise measurement of the Hubble constant from a single lensed SN.  The SN is in an evolved galaxy at $z=1.95$, gravitationally lensed by the massive foreground galaxy cluster MACS J0138.0-2155.  This is the most distant multiply-imaged supernova yet discovered.  We classify the supernova as a Type Ia with {$>90\%$} probability based on archival imaging observations from the \textit{Hubble Space Telescope}.  Three lensed images of the SN are detected with relative time delays of $<$200 days, and we predict that a fourth image in the component closest to the cluster core will appear in the year 2037$\pm$2.  Observing the light curve of that future image could provide a time delay precision of $\approx 7$ days over an extraordinary 20 year baseline.

@@

\section*{Main Text:}

@@

One promising cosmological tool uses %time delays of 
gravitational lens systems in which a background
source appears as multiple images that arrive to the observer with relative delays \cite{einstein_lens_1936}.
%, which %for a given lens potential 
When such a strongly-lensed source is variable, one can measure the time delay between any pair of lensed images and derive a ratio of angular diameter distances to the foreground lens and background source. 
This distance ratio is sensitive to cosmological parameters, such as 
%the expansion rate of the universe, 
$H_0$  and the dark energy equation-of-state, $w$ \cite{refsdal_possibility_1964,coe_cosmological_2009,linder_lensing_2011}.
In recent years this method has been successfully applied to lensed
quasars \cite{suyu_cosmology_2014,bonvin_cosmograil_2019}, with six high-precision measurements to date \cite{wong_h0licow_2019}. As this sample of time delay lenses grows to several dozen, it is expected to deliver a measurement of $H_0$ with 1\% precision \cite{suyu_cosmological_2018}.

Gravitationally lensed SN with multiple images present an attractive new addition to this field. Most notably, SN exhibit relatively simple photometric behavior, with well-understood light curve shapes and colors---in contrast to the stochastic variation of quasars.  However, despite the promise of lensed SN time delay cosmography as a cosmological distance probe, to date, there have been only two lensed SN observed with multiple images \cite{kelly_multiple_2015,goobar_iptf16geu:_2017}. The first, {\it SN Refsdal}, was a peculiar Type II SN whose image with the longest delay was missed \cite{kelly_sn_2016}. The second, {\it SN 2016geu}, was a Type Ia SN with short delays that make high precision time delay measurements impossible \cite{dhawan_magnification_2019}. Here we present the discovery 
of a third lensed SN resolved into multiple images: \SNABC.

We discovered \SNABC ({\it AstroNotes citation})
using data from the Hubble Space Telescope (HST) program {\it REsolved QUIEscent Magnified Galaxies} (REQUIEM, HST-GO-15663, PI:Akhshik). This project targets rare examples of massive galaxies with low specific star formation rates that have been magnified by strong gravitational lensing. %These are typically too faint and compact to be resolved even with Hubble, but due to lensing are magnified. This provides a unique opportunity to  understand the still mysterious quenching mechnism. 
The brightest and most spectacular galaxy targeted by REQUIEM
is MRG-M0138, first discovered as a massive red galaxy (MRG) at $z=1.95$ \cite{newman_resolving_2018} behind the massive galaxy cluster MACS J0138.0-2155 \cite{ebeling_macs_2001}.
MRG-M0138 is quadruply lensed by a foreground galaxy cluster at $z=0.338$.  
During analysis of REQUIEM observations obtained 13--14 July 2019 \cite{materials_methods_2020} we discovered three point 
sources that were present in archival HST images from 18--19 July 2016, part of the program 
that first confirmed the MRG-M0138 galaxy as a strongly-lensed object (HST-GO-14496; PI:Newman). 
Each point source is within 5 arcseconds of one of the four MRG-M0138 images.  None of the
three point sources are present in the REQUIEM HST data in 2019 (Fig. \ref{fig:layout}). We infer that 
the point sources are multiple images of a single astrophysical 
transient in MRG-M0138, most likely a SN \cite{materials_methods_2020}.  
%The official designation for this multiply-imaged astrophysical transient is \TNSname. 
%Formal designation as a SN requires spectroscopic confirmation, but with a high-confidence photometric classification we coin the name \SNABC for this unusual source.

To construct a lens model for the MACS J0138.0-2155 cluster we use the \lenstool software \cite{jullo_bayesian_2007,kneib_lenstool_2011} to model the mass distribution in the cluster core as the combination of a cluster-scale and multiple galaxy-scale potentials \cite{materials_methods_2020}.  To avoid unintended bias, we kept the lens model development completely separate from the analysis of the SN.  Only upon independent completion of both were the results combined for the analysis described here.  The input model constraints are the positions and redshifts of the MRG-M0138 galaxy at $z=1.95$ (both the central light peak and the SN location in each image) as well as a multiply-imaged background 
galaxy at $z=0.766$, both having secure spectroscopic redshifts \cite{materials_methods_2020}.  
From this model we derive estimates for the lensing magnification and time delay of each of the SN images, including the predicted fourth image (Table \ref{tab:time_delays}).
The lensing model predicts that the SN should appear in the fourth MRG-M0138 image in the year 2037$\pm$1.5, demagnified with $\mu=0.4\pm0.2$. The predicted location for this future image is $\alpha=01:38:04.15,\delta=-21:55:24.659$.
A fifth image will also appear at date later still, located near the center of the cluster and much more significantly demagnified, so it will not be easily observable.

To realize the cosmological promise of \SNABC 
we need to estimate the age of each SN image, which in turn constrains the lensing time delays. For this
goal, it is necessary to have a firm determination of the transient’s class. Expected time delays and magnifications from the lens model exclude any of the various rapidly evolving and low-luminosity stellar transient classes, strongly suggesting that it is a SN. The first-order SN distinction remaining is between a Type Ia SN---the explosion of a white dwarf star in a binary system---and a core collapse SN (CCSN)---the end-point of a star with mass $>10 M_{\odot}$. 
The properties of the host galaxy can inform this classification because CCSN are limited to galaxies with young stellar populations. Limits on the specific star formation rate and age for this host, MRG-M0138, show it to be a massive but very quiescent and evolved galaxy, %($\log(M_*/M_\odot)=11.7\pm0.2; )$ , 
unlikely to retain any significant population of high-mass stars \cite{newman_resolving_2018}. Based on observed properties of the host galaxy alone, we find a 62-75\% probability that \SNABC is a Type Ia SN \cite{materials_methods_2020}. Adopting the lens model magnifications for the three observed SN images (Table \ref{tab:time_delays}) we can locate each SN image in color-magnitude space (Fig. \ref{fig:class}). This provides more classification leverage, yielding p(Ia)=95\% \cite{materials_methods_2020}. By also including the model-predicted time delays, we can treat the three SN images effectively as three points on a common SN light curve, and we find p(Ia)=94\% \cite{materials_methods_2020}.
An improved classification can be achieved with spectroscopy and multi-band photometry upon arrival of the fourth image.

The color of a SNIa evolves substantially over its lifetime as the photosphere expands and cools, revealing different layers of the expanding shell and driving episodes of recombination \cite{woosley_type_2007,kasen_type_2009}. Since the phenomenon of gravitational lensing in general is achromatic, this color evolution makes it possible to derive an age constraint that is largely independent of the lens model \cite{materials_methods_2020}.  Combining this information with magnification constraints from the lens modeling helps break parameter degeneracies, yielding the measured delays in Table \ref{tab:time_delays} \cite{materials_methods_2020}. 
Remarkably, the ages of image 1 and 3 are constrained to better than $\pm$20 days, despite having only a single epoch of photometric data.    These uncertainties may be 
further reduced when the future fourth image is observed with high-precision, multi-epoch photometry.  Such a light curve will pin down the intrinsic SN light curve parameters that are shared by all images, and break remaining parameter degeneracies (see supplemental materials).

Future large-scale surveys such as the Vera C. Rubin Observatory 
and the Nancy Grace Roman Space Telescope will observe dozens to hundreds of lensed SN over their mission lifetimes \cite{oguri_gravitationally_2010,goldstein_rates_2019,wojtak_magnified_2019}. However, many of these lensed SN will require substantial follow-up resources to be made cosmologically useful. Furthermore, almost all will have significantly shorter 
delays (of order 10--100 days) and therefore higher relative uncertainties than \SNABC.  Since it is the {\it fractional} time delay uncertainty that 
propagates through to any time delay distance measurement, the extraordinarily long time delays of cluster-lensed SNe like SN Requiem can deliver significantly better time delay precision, with comparable observational cost.  In fact, the long time baseline of lensed SN like \SNABC effectively insures that their cosmological precision is not limited by the time delay measurement errors.

In Figure \ref{fig:cosmo} we illustrate the cosmological value of time delay cosmography with objects like \SNABC  by considering a small sample of 5 ``\reqlike'' SN.  For a comparison on equal footing, we evaluate equal-sized lensed SN samples 
that mimic those expected from the Rubin Observatory and  the Roman Space Telescope.  Differences in the 
ratio of source to lens redshift account for the different cosmological parameter covariances, reflected in the orientations of the contours in these plots. Because of the unusually large redshift ratio of \SNABC, the constraints 
from the \reqlike sample are nicely complementary to the low-$z$ Rubin Observatory and high-$z$ Roman Space Telescope samples. 
Furthermore, when we assume that the lens modeling uncertainty is equal for each sample (amounting to a 5\% uncertainty in each time delay distance), the \reqlike sample has leverage that is comparable to the same number of lensed SN from the wide-field surveys. However, if we improve  the lensing constraints to 2\% precision, the \reqlike contours tighten dramatically, reflecting the fact that these long-time-baseline events are limited only by the lens modeling uncertainty.

To achieve these cosmological constraints in practice, such a sample could be developed with regular monitoring of cluster-scale lenses, partnered with modest follow-up to characterize any lensed SN discovered.  As \SNABC shows, even a single epoch of photometric data at the time of detection may be sufficient to provide the necessary anchor point for this promising cosmological tool.  %After that, all we need is patience.

HST observations enabled us to find this SN.  We anticipate that HST may be de-orbited and make its final plummet to Earth around the time of the reappearance of \TNSname, so we coin the name {\it SN  Requiem} as an ode to the vast new discovery space that HST continues to unveil.

@@

\clearpage
\begin{figure*}
    \centering
    \includegraphics[draft=False,width=0.9\textwidth]{Paper/Figures/fig1_layout_sn4.pdf}
    \caption{Overview of the MACSJ0138 cluster field and the locations of the \SNABC images  (SN1--3) and its multiply-imaged host galaxy (H1--4). The wide-field view in $a)$ is 40\arcsec\ on a side with ticks indicating 10\arcsec\ intervals.  
    %Labels marked 3.1--3.4 indicate the locations of a separate  multiply-imaged [O\textsc{ii}]-emitter at $z=0.77$ used to help constrain the lensing potential. 
    Panels \emph{b}--\emph{i} show 4\arcsec\ cutouts around the lensed SN images with 1\arcsec\ ticks.  The three-color images are generated from the WFC3/IR filters as indicated; panels \emph{bcde} show the imaging from 2016 July where the SN was visible and \emph{fghi} show the later imaging from 2019 July where the the SN has faded away.  Panels \emph{ei} indicate the location of the fourth image predicted to appear in $\approx$2037. All panels use the late-epoch F125W imaging for the green channel; nevertheless, it is immediately clear that the SN2 image is substantially bluer than the other two, which helps to constrain the relative age of each SN image and the transient classification as a likely Type Ia supernova explosion.}

    \label{fig:layout}
\end{figure*}
\clearpage
\begin{figure}
  \centering
    \includegraphics[draft=False, width=\textwidth]{Paper/Figures/classification_contours_timeline.pdf}
    \caption{Classification information for SN Requiem based on its position in color-magnitude space. \emph{Left:} Contours show the population distributions for normal SNe of Type Ia (red), Type Ib/Ic (gold), and Type II (green), drawn to enclose 68\% (interior solid lines) and 95\% (exterior shaded regions) of each SN population.  Each SN sub-class was simulated at $z=1.95$, and samples from their expected light curves were drawn uniformly in time. Open markers show the observed photometry of the SN. Dotted vertical lines mark the lensing magnification correction, terminating in closed markers and solid bars that indicate the magnification in our preferred lens model and its systematic uncertainty, respectively.  \emph{Right:} The evolution of a typical Type Ia SN at $z=1.95$ in observed color-brightness space is shown by a colored line, with the line color indicating SN age in observer-frame days relative to peak brightness.  White diamonds correspond to the times labeled on the colorbar at right. Grey shading shows the typical range of luminosities observed for the Type Ia SN population in the nearby universe \cite{wang_determination_2006}. 
    %An interactive 3-D version of this figure is available online at \href{https://plot.ly/~jpierel/6}{https://plot.ly/~jpierel/6}
    \label{fig:class}
    }
\end{figure}

\clearpage
\begin{figure}
    \centering
    \includegraphics[draft=False,trim={2cm, 3cm, 3cm, 4cm},clip,width=\textwidth]{Paper/Figures/snrequiem_hw_w0wa_apples_to_lsst_ngrst_4panel.pdf}
    \caption{  
    The sensitivity of \SNABC to the expansion rate of the universe ($H_0$) and dark energy equation-of-state parameters ($w_0, w_a$). Dashed lines and the white marker show the input values (upper right of the top panels) that were used for this simulation. Contours show the 1$\sigma$ confidence region for a projected sample of 5 lensed SNIa time delays from the Rubin Observatory (blue) and the Roman Space Telescope (red) compared to 5 events like \SNABC (black).  The left and right columns assume a 5\% and 2\% contribution to the distance uncertainty from modeling the lens potential, respectively. 
    In order to isolate the plotted parameters, all contours are constructed assuming perfect knowledge of the cosmological parameters that are not shown. 
    }
    \label{fig:cosmo}
\end{figure}
